Uncovering underlying musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and aural music skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks

### 2023-04-26 figures/equations/ReLU.tex
Updates from Overleaf

The Rectified Linear Unit (ReLU) activation function is a simple yet highly effective non-linear function used in neural networks. It helps introduce non-linearity to the model and accelerates training due to its computational efficiency. Here's a brief LaTeX explanation of the ReLU function:

ReLU is a piecewise linear function that outputs the input value if it's positive and zero otherwise. Mathematically, it can be represented as:

@@

Drawbacks: Despite its benefits, ReLU has some drawbacks, such as the "dying ReLU" issue, where neurons can become inactive during training and never fire again. This problem can be addressed using variants like Leaky ReLU or Parametric ReLU.

>>>>

The Rectified Linear Unit (ReLU) activation function is a simple yet highly effective non-linear function used in neural networks. It helps introduce non-linearity to the model and accelerates training due to its computational efficiency.

ReLU is a piecewise linear function that outputs the input value if positive and zero otherwise. Mathematically, it can be represented as:

@@

Drawbacks: Despite its benefits, ReLU has some flaws, such as the "dying ReLU" issue, where neurons can become inactive during training and never fire again.

### 2023-04-26 introduction/Subsections/Motivation.tex
Updates from Overleaf

\section{Motivation}
As a music lover, professional musician, and connoisseur, I am interested in gaining a deeper understanding of the fundamental building of musical composition, the various techniques used to create musical works, and capturing underlying relationships among them. I am deeply interested in learning how to retrieve useful embedded information within music.

@@

I have consistently been inspired by musicians and researchers who strive to uncover musical ground truth within the existing tonal paradigm, such as Heinrich Schenker, or by challenging it, as exemplified by George Russell \cite{LydianRussell}. Both approaches aim to identify abstract concepts that reinforce or disrupt the tonal foundation. Their ultimate goal is to advance the tonal landscape, providing musicians with a dependable playground for growth, development, and further understanding tonal paradigms.

@@

\begin{figure}[ht]
\includegraphics[clip,width=\columnwidth]{figures/schenkerian analysis/SchubertOp4no3.png}% 
\caption{Small excerpt of \textit{Wandrers Nachtlied, Op. 4, D. 224} by Franz Schubert. We can see the passage's original score, the schenkerian unfolding of the melody, the chord degrees, and the tonal function.}
\label{fig:Wandrers Nachtlied, Op. 4, D. 224}
\end{figure}

>>>>

\section{Motivation}

This thesis is driven by personal, academic, and industrial motivations and is conducted in collaboration with Epidemic Sound \cite{EpidemicSite}, a prominent organization in the music technology sector. A deeper understanding of music and its fundamental structures fulfills intellectual curiosity and generates potential advancements and benefits for the company's technical products and services.

\subsubsection{Personal motivation}

As a music enthusiast, professional musician, and scholar, I am passionately committed to exploring the foundational elements of musical composition, investigating the diverse techniques employed in creating musical works, and deciphering the intricate relationships among them. This passion fuels my quest for knowledge on extracting valuable information embedded within music, contributing to developing more sophisticated and effective music-related technologies and solutions in the industrial landscape.

I advocate for exposure as the primary method of learning music, emphasizing the importance of immersing oneself in various musical experiences to foster a holistic understanding of music as an art form. Individuals can cultivate well-rounded musical expertise by engaging with different styles, genres, techniques, and learning approaches, ultimately promoting creativity and experimentation in both personal and professional domains.

\subsubsection{Academic motivation}

The emergence of self-supervised models that learn embedding spaces for retrieving musical content from audio signals has unveiled new opportunities in academic research. By capturing high-level semantic information from raw audio data in an unsupervised manner, these models can potentially revolutionize various aspects of the music industry. Their applications encompass music information retrieval, sharing learned latent representations for MIR \cite{HamelTransferSimilarity}, and fostering interdisciplinary research and innovation in musicology, computer science, and artificial intelligence.

\subsubsection{Industrial and corporate motivation}

In the industrial and corporate realm, the embedding spaces generated by these models can be utilized to create personalized and intelligent music recommendation systems \cite{Chen2020LearningRecommendation}\cite{epidemic}, facilitate effective enforcement of intellectual property rights, and contribute to the development of more sophisticated audio editing-production tools and technical products \cite{WonEmotionStories}. These advancements ultimately enhance our interaction with music in the digital age, promoting innovation and collaboration across the music industry.

@@

I have consistently been inspired by musicians and researchers who strive to uncover musical ground truth within the existing tonal paradigm, such as Heinrich Schenker, or by challenging it, exemplified by George Russell \cite{LydianRussell}. Both approaches aim to identify abstract concepts that reinforce or disrupt the tonal foundation. Their ultimate goal is to advance the tonal landscape, providing musicians with a dependable playground for growth, development, and further understanding of tonal paradigms.

@@

\begin{figure}[ht]
\includegraphics[clip,width=\columnwidth]{figures/schenkerian analysis/SchubertOp4no3.png}% 
\caption{Small excerpt of \textit{Wandrers Nachtlied, Op. 4, D. 224} by Franz Schubert. We can see the passage's original score, the schenkerian unfolding of the melody, the chord degrees analysis, and the tonal function.}
\label{fig:Wandrers Nachtlied, Op. 4, D. 224}
\end{figure}

### 2023-04-26 introduction/Subsections/Objectives.tex
Updates from Overleaf

This work aims to uncover a generalization of the fundamental tonal-working structure of western-tradition music by learning and embedding space used to facilitate its application in downstream MIR tasks.

In recent years, using neural networks has enabled low-dimensional musical embeddings that capture crucial musical features and allow the computation of content-based similarity between musical passages \cite{Chen2020LearningRecommendation}\cite{KimLEARNINGLOSS}\cite{Hung2022Feature-informedClassification}\cite{WonEmotionStories}. By acquiring an embedding function $f(x) \in \mathbb{R}^d$ that maps a given course $x$ to a vector in the real space $\mathbb{R}^d$, we can measure the similarity between two musical passages using the Euclidean distance, as represented by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The effectiveness of such systems depends on the level of musical abstraction and the embedding space's ability to interpret music in a way that aligns with human perception. Therefore, a suitable embedding space is crucial to ensure the output aligns with human expectations.

This study aims to comprehensively analyze musical works from diverse styles and genres by incorporating computational techniques from Music Information Retrieval (MIR) and a musicological perspective to contextualize the findings. Aural skills are required to comprehend musical works through listening, as they involve perceptual and cognitive abilities. The underlying musical structure refers to the systematic arrangement of elements that produce a coherent and expressive musical work. This study aims to offer fresh insights into these components of musical composition and advance our understanding of their relationships.

Let's denote the single embedding for a music piece as $e(M)$.

Given a music piece $M$, we aim to learn a single embedding, $e(M)$, simultaneously capturing the underlying content (melody, harmony, critical, holistic elements) and the production aspects. The embedding space should be structured so that similar content and production features are clustered together, effectively isolating the two elements. Formally, our objective is to learn the embedding function $e(M)$ such that:

\begin{itemize}
  \item Similar content features are closer in the embedding space: $\|e(M_1) - e(M_2)\| < \epsilon_c$ for music pieces $M_1$ and $M_2$ with similar content.
  \item Similar production features are closer in the embedding space: $\|e(M_1) - e(M_3)\| < \epsilon_p$ for music pieces $M_1$ and $M_3$ with similar production.
  \item Distinct content and production features are further apart in the embedding space.
\end{itemize}

We describe learning a single embedding $e(M)$ that captures content and production features. The embedding space is structured to cluster similar content and production features together, effectively isolating the two aspects.

Music is a complex and multi-faceted phenomenon, and implementing a model mimicking human aural music skills is challenging. However, the latest deep learning techniques provide a promising avenue to build a model that achieves good results for a specific task in music. Therefore, this study will utilize deep learning techniques to investigate the correlation between aural skills and the underlying tonal-working structure of music.

Furthermore, this study will argue that MIR research should shift from a purely technical-centered approach to a more balanced and comprehensive one incorporating other relevant domains such as musicology and music theory. By doing so, we can better understand the interplay between technical and musical aspects of music and improve our understanding of music signals and the music itself.

\subsection{Applications}
%%%%%%%%%%%%%%%%%%%%%%%%
%POTENTIAL APPLICATIONS
%%%%%%%%%%%%%%%%%%%%%%%%%

A self-supervised model that learns an embedding space to retrieve musical content from raw audio signals could have several potential applications and downstream tasks in MIR. Here are some examples:

>>>>

This study seeks to discover a generalization of the fundamental tonal structure in Western music by creating and examining an embedding space that can be applied to downstream Music Information Retrieval (MIR) tasks.

Neural networks have recently facilitated the development of low-dimensional musical embeddings that capture essential musical features, enabling the computation of content-based similarity between musical passages \cite{KimLEARNINGLOSS}\cite{Hung2022Feature-informedClassification}. An embedding function $f(x) \in \mathbb{R}^d$ maps a given course $x$ to a vector in the real space $\mathbb{R}^d$, allowing the measurement of similarity between two musical passages using Euclidean distance, as denoted by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The success of such systems hinges on the level of musical abstraction and the embedding space's capacity to interpret music in a manner consistent with human perception. As a result, a suitable embedding space is vital for producing output that aligns with human expectations.

This research aims to thoroughly analyze musical works from various styles and genres, employing computational techniques from Music Information Retrieval (MIR) and a musicological perspective to contextualize the findings. Aural skills, encompassing perceptual and cognitive abilities, are necessary for understanding musical works through listening. The underlying musical structure consists of a systematic arrangement of elements that generate a coherent and expressive musical work. This study seeks to provide new insights into these components of musical composition and further our comprehension of their interrelationships.

We denote a single embedding for a music piece as $e(M)$.

For a given music piece $M$, we aim to learn a single embedding, $e(M)$, that captures both the underlying content (melody, harmony, critical, holistic elements) and production aspects. The embedding space should be structured to cluster similar content and production features together, effectively separating the two aspects. Formally, our objective is to learn the embedding function $e(M)$ such that:

\begin{itemize}
\item Similar content features are closer in the embedding space: $|e(M_1) - e(M_2)| < \epsilon_c$ for music pieces $M_1$ and $M_2$ with similar content.
\item Similar production features are closer in the embedding space: $|e(M_1) - e(M_3)| < \epsilon_p$ for music pieces $M_1$ and $M_3$ with similar production.
\item Distinct content and production features are further apart in the embedding space.
\end{itemize}

We propose learning a single embedding $e(M)$ that captures content and production features, with the embedding space structured to cluster similar aspects together, effectively distinguishing between the two.

Music is a complex and multi-dimensional phenomenon, and developing a model that emulates human aural skills is challenging. Nonetheless, recent advancements in deep learning offer promising avenues for creating a model that achieves satisfactory results in specific music-related tasks. This study will employ deep learning techniques to explore the correlation between aural skills and the underlying tonal structure of music.

Additionally, this research contends that MIR studies should transition from a solely technical-centered approach to a more balanced and comprehensive method that integrates other pertinent domains such as musicology and music theory. This shift will enable a better understanding of the interplay between technical and musical aspects of music, enhancing our comprehension of music signals and music itself.

### 2023-04-26 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

\section{Model architecture and training strategy}
Piaget's theory of cognitive development \cite{Huitt2003PiagetsDevelopment} posits that children acquire knowledge from sensory and motor experiences in their early stages, from birth to around 18 months. This period called the \textit{sensorimotor stage}, sees the emergence of early representational thought through basic actions such as sucking, grasping, looking, and listening. As children progress through different developmental stages until adolescence and adulthood, their reasoning progressively moves toward abstract ideas and deductive logic. Schemas, higher-order cognitive structures hypothesized to underlie many aspects of human knowledge and skill, emerge during this process, which Piaget interprets through an equilibration mechanism that balances new information with old knowledge. This mechanism involves \textit{assimilation} and \textit{accommodation}, which entail taking in further information to fit pre-existing schemas and modifying existing schemas resulting from the latest news. In the context of learning, complex structures are based on simpler ones, and knowledge transfer occurs when dynamic systems capable of generalization are created. \cite{audioselfsupsurvey}

@@

Contrastive Self-Supervised Learning of Musical Representations (CoSSLeMuR) is an approach that combines the principles of Self-Supervised Learning (SSL) \cite{audioselfsupsurvey} and Contrastive Learning of Musical Representations (CLMR) \cite{CLMR2021} to learn discriminative and valuable representations of music without relying on explicit labels. By leveraging the structure and content of music and contrasting different augmentations or transformations of the same musical piece against randomly (or not) selected pieces, CSSLMR models can effectively capture the underlying structure and known invariance in the data.

@@

where:

@@

    Let $A$, $P$, and $N$ represent the anchor, positive, and negative samples, respectively. The maximum length in the batch is given by:

@@

    The hardest negative index for each anchor-positive pair is determined by maximizing the difference between the anchor-negative and anchor-positive distances:

@@

If the model is "supervised", dropout with a rate of 0.5 is applied.

@@

We have changed and adapted the original architecture for our experiments to our purpose and needs—the weights initialization and the final average pooling step.

@@

The final positive waveform $W_p$ is obtained with noise added according to the chosen SNR. This pipeline enhances the diversity of the positive samples and aims to improve the model's ability to retrieve similar-sounding samples sharing the same underlying musical language.

@@

\begin{equation}
L_a = \text{shape}(a)
\end{equation}
\begin{equation}
T_a = \frac{L_a}{f_s}
\end{equation}

where $\text{shape}(a)$ returns the shape of the audio clip, and $f_s$ is the sample rate of the audio.

@@

\begin{equation}
c_i = a[..., s_i : s_i + L_{c,i}]
\end{equation}
\begin{equation}
S_c = \text{cumsum}([0] + [L_{c,1},\ldots,L_{c,n_c-1}])
\end{equation}
\begin{equation}
s_i = S_c[i]
\end{equation}
\begin{equation}
i=1,\ldots,n_c
\end{equation}

\item Shuffle the resulting list of chunks randomly to obtain the reordered chunks $c'_1,\ldots,c'_{n_c}$:

\begin{equation}
c' = \text{shuffle}([c_1,\ldots,c_{n_c}])
\end{equation}

\item Concatenate the shuffled chunks $c'_1,\ldots,c'_{n_c}$ to create the negative clip $n$:

\begin{equation}
n = \text{concatenate}(c')
\end{equation}

@@

The triplet loss aims to ensure that the distance between an anchor and a positive sample (both from the same class) is smaller than the distance between the anchor and a negative sample (from a different class) by a margin.

$\mathcal{L} = \frac{1}{N}\sum_{i=1}^{N} \max \left( d\left(a_i, p_i\right) - d\left(a_i, n_i\right) + m, 0 \right)$

In this equation:

$N$ is the number of triplets in the batch
$a_i$ is the $i$-th anchor
$p_i$ is the $i$-th positive sample
$n_i$ is the $i$-th negative sample
$d(\cdot, \cdot)$ is the Euclidean distance function
$m$ is the margin; commonly set to 0.2 by default
The forward method computes the triplet loss as follows:

Calculate the pairwise Euclidean distance between the anchor and positive samples, denoted as $d(a_i, p_i)$.
Calculate the pairwise Euclidean distance between the anchor and negative samples, denoted as $d(a_i, n_i)$.
Compute the loss for each triplet in the batch by subtracting the negative distance from the positive distance and adding the margin: $d(a_i, p_i) - d(a_i, n_i) + m$.
Apply the ReLU function, which is equivalent to taking the maximum value between the computed value and 0: $\max \left( d(a_i, p_i) - d(a_i, n_i) + m, 0 \right)$.
Calculate the mean of the losses across all triplets in the batch.

>>>>

\section{Model architecture and training strategy}
Piaget's theory of cognitive development \cite{Huitt2003PiagetsDevelopment} posits that children acquire knowledge from sensory and motor experiences in their early stages, from birth to around 18 months, during the \textit{sensorimotor stage}. This stage sees the emergence of early representational thought through basic actions such as sucking, grasping, looking, and listening. As children progress through different developmental stages until adolescence and adulthood, their reasoning progressively moves toward abstract ideas and deductive logic. Schemas, higher-order cognitive structures hypothesized to underlie many aspects of human knowledge and skill, emerge during this process, which Piaget interprets through an equilibration mechanism that balances new information with old knowledge. This mechanism involves \textit{assimilation} and \textit{accommodation}, which entail taking in further information to fit pre-existing schemas and modifying existing schemas resulting from the latest information. In the context of learning, complex structures are based on simpler ones, and knowledge transfer occurs when dynamic systems capable of generalization are created. \cite{audioselfsupsurvey}

@@

Contrastive Self-Supervised Learning of Musical Representations (CSSLMR) is an approach that combines the principles of Self-Supervised Learning (SSL) \cite{audioselfsupsurvey} and Contrastive Learning of Musical Representations (CLMR) \cite{CLMR2021} to learn discriminative and valuable representations of music without relying on explicit labels. By leveraging the structure and content of music and contrasting different augmentations or transformations of the same musical piece against randomly (or not) selected pieces, CSSLMR models can effectively capture the underlying structure and known invariance in the data.

@@

Where:

@@

    Let $A$, $P$, and $N$ represent the anchor, positive, and negative samples. The maximum length in the batch is given by:

@@

    The "hardest" negative index for each anchor-positive pair is determined by maximizing the difference between the anchor-negative and anchor-positive distances:

@@

If the model is "supervised", dropout with a rate of 0.5 is applied; a regularization technique randomly drops out neurons during training to prevent overfitting, with a 0.5 rate meaning half of the neurons are dropped out.

@@

We have adapted the original architecture for our experiments to our purpose and needs—the weights initialization and the final average pooling step.

@@

The final positive waveform $W_p$ is obtained with noise added according to the chosen SNR. This pipeline enhances the diversity of the positive samples and aims to improve the model's ability to retrieve content-based similar-sounding samples sharing the same underlying musical language.

@@

\begin{equation}
L_a = \text{len}(a)
\end{equation}

\begin{equation}
T_a = \frac{L_a}{f_s}
\end{equation}

where $\text{len}(a)$ returns the shape of the audio clip, and $f_s$ is the sample rate of the audio.

@@

\begin{equation}
c_i = a[..., s_i : s_i + L_{c,i}]
\end{equation}
\begin{equation}
S_c = [0] + [L_{c,1},...,L_{c,n_c-1}]
S_c = \sum_{i=0}^{n_c-1} S_{c,i}
\end{equation}
\begin{equation}
s_i = S_c[i]
\end{equation}
\begin{equation}
i=1,\ldots,n_c
\end{equation}

\item Shuffle the resulting list of chunks randomly to obtain the reordered chips $c'_1,\ldots,c'_{n_c}$:

Let $p$ be a random permutation of the indices $1$ to $n_c$. Then, the shuffled array $c'$ is given by:
\begin{equation}
c' = [c_{p(1)}, c_{p(2)}, \dots, c_{p(n_c)}]
\end{equation}

\item Let $c'_1, \dots, c'_{n_c}$ be the shuffled chunks. Then, the negative clip $n$ is obtained by concatenating the shuffled chunks:
\begin{equation}
n = [c'_1, \dots, c'_{n_c}]
\end{equation}

@@

The mathematical expression for the triplet loss function with a variable margin can be written as follows:

\begin{equation}
L(\mathbf{a}, \mathbf{p}, \mathbf{n}) = \frac{1}{N} \sum_{i=1}^{N} \max \left(0, m + \left| \mathbf{a}_i - \mathbf{p}_i \right|_2^2 - \left| \mathbf{a}_i - \mathbf{n}_i \right|_2^2 \right)
\end{equation}

In the triplet loss function, $d(\mathbf{a}, \mathbf{p})$ and $d(\mathbf{a}, \mathbf{n})$ denote the Euclidean distance (L2-norm) between the anchor $\mathbf{a}$ and the positive sample $\mathbf{p}$ and the negative sample $\mathbf{n}$, respectively. The margin $m$ is a positive constant determining the minimum distance between the anchor-positive and anchor-negative pairs.

The $\max(0, ...)$ function computes the maximum of the expression within the parentheses and 0. This function aims to ensure that the loss is only calculated for triplets that violate the margin constraint, where the negative sample is closer to the anchor than the positive sample by at least the margin. When the distance between the positive and negative samples is greater than the margin, the loss is zero. Otherwise, the loss is proportional to the extent to which the constraint is violated.

The final form of the loss function is obtained by averaging the loss over a mini-batch of triplets.

\begin{equation}
L = \frac{1}{N} \sum_{i=1}^{N} L(anchor_i, positive_i, negative_i)
\end{equation}

Here, $N$ is the number of triplets in the mini-batch, and $L(anchor_i, positive_i, negative_i)$ represents the loss function computed for the i-th triplet in the batch.
